Viscous Erosion in Porous Media

### 2018-03-03 docs/methods/erosion_methods.tex
Adding comments for intro section.

Because of the length and velocity scales of groundwater flow, we
restrict our attention to small Reynolds number flows.  For groundwater
flow, Darcy's law is often used, but this does not resolve the pore
structure and assumes a homogeneous geometry which is often not valid.
Instead, we use the incompressible Stokes equations to represent the
flow, and the individual pores are resolved which results in a complex
multiply-connected geometry.  A boundary integral equation is
well-suited for solving the incompressible Stokes equations in a complex
geometry since it can resolve individual grains and the non-negligible
interactions between bodies are resolved. Moreover, high-order accuracy
and computational efficiency are obtained. To link the eroding body and
fluid dynamics, the body is eroded in the normal direction at a rate
that is proportional to the magnitude of the shear
stress~\cite{ris-moo-chi-she-zha2012}.  Tracking the interface of
fluid-structure interaction problems with a long time horizon is
particularly---the mesh is often distorted and tangled by the dynamics
and regions of high curvature develop.  These two issues are resolved
by, first, introducing a tangential velocity field to maintain a grid
that is equispaced in arc length, but that does not affect the dynamics,
and, second, a curvature penalization term.  Furthermore, the frame of
reference is changed to \thL~variables so that the curvature term
appears as a diffusive term that can easily be treated implicitly
to avoid stiffness.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\paragraph{Contributions} The present study develops numerical methods
to simulate erosion of multiply-connected two-dimensional geometries.
Since groundwater flow occurs at small spatial and velocity scales, we
assume the flow is viscous (ie.~zero Reynolds number).  By allowing for
complex multiply-connected domains, we capture the effect that
interacting bodies have on erosion, and the grains in the flow are
individually resolved rather than homogenized.  We start by validating
our method on a single circular body placed symmetrically in a pipe
flow.  By artificially fixing the area of the body as a function of
time, we are able to characterize the long-time behavior of the shape
and the resulting flow and compare these results with analytical
results.  Then, we compute the vorticity of erosion in
multiply-connected geometries to better characterize erosion in porous
media.

@@

A challenge with erosion is that it creates corners at points on the
body where the shear stress vanishes.  While recent work does allow for
integral equation methods to resolve corners, we employ a different
method.  First, we add a regularization term to the shear stress to
penalize regions of high curvature.  Second, we apply a Gaussian filter
to smooth the absolute value of the shear stress before the body shape
is updated.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\paragraph{Limitations} There are two main limitations of the current
work.  First, the flow is assumed to be two-dimensional.  However, by
assuming that the flow is two-dimensional, we are able to simulate the
erosion of more bodies and with higher resolution than would be
available in three dimensions.  Moreover, several flows can be described
to good accuracy under a two-dimensional approximation, but certain
phenomena occur only in three dimensions.  Extending the current work to
three dimensions is possible with a boundary integral equation
formulation.  In particular, the fluid solvers exist~\cite{mal-bir2015},
and the main challenges of a BIE solver have been addressed by other
groups including surface representation~\cite{yin-bir-zor2006,
vee-rah-bir-zor2011}, near-singular integration~\cite{yin-bir-zor2006,
kli-tor2016b}, and fast summation methods~\cite{yin-bir-zor2004,
ros-ols2015}.  The more significant challenge of extending our work to
three dimensions is guaranteeing that the mesh does not become
distorted, which can be done, for example, by using an extension of the
two-dimensional {\thL} formulation~\cite{amb-sie-tlu2013}.

@@

In this section, we describe the boundary integral equations employed for the fluid
equations \cite{Moore2007}
\todo[inline]{Bryan do you want to add a bunch of references to your boundary integral papers here (or somewhere else)?}
In particular, we solve the incompressible Stokes equations
with Dirichlet boundary conditions 
\begin{alignat*}{3}
  \mu\Delta \uu &= \grad p, \qquad && \xx \in \Omega, \\
  \grad \cdot \uu &= 0,   && \xx \in \Omega, \\
  \uu &= \ff,  && \xx \in \bd\Omega.
\end{alignat*}
The boundary condition $\ff$ is the prescribed velocity $\UU$ on
$\Gamma$, and it is zero on the grains $\gamma$.  As the viscosity $\mu$
also sets the time scale, we take $\mu=1$.  We start by defining the
double-layer potential
\begin{align}
  \DD[\eeta](\xx) = \frac{1}{\pi} \int_{\bd\Omega} 
    \frac{\rr \cdot \nn}{\rho^2} \frac{\rr \otimes \rr}{\rho^2} 
    \eeta(\yy) \, ds_\yy, \quad \xx \in \Omega,
    \label{eqn:velocityDLP}
\end{align}
where $\rr = \xx - \yy$, $\rho = \|\rr\|$, and $\eeta$ is an unknown density function.  The
double-layer potential cannot capture velocity fields corresponding to
rigid body motions~\cite{pow-mir1987}.  The rank deficiency is removed
by complementing the double-layer potential with a Stokeslets and
rotlets centered inside each interior body
\begin{align}
  S[\llambda_\ell,\cc_\ell] = \frac{1}{4\pi} \left(-\log \rho \II + 
    \frac{\rr \otimes \rr}{\rho^2} \right)\llambda_\ell
  \quad \text{and} \quad
  R[\xi_\ell,\cc_\ell] = \frac{\rr^\perp}{\rho^2}\xi_\ell, 
  \quad \ell=1,\ldots,M,
  \label{eqn:stokeslet_rotlet}
\end{align}
where  $\rr = \xx - \cc_\ell$, $\cc_\ell$ is a point inside the
$\ell^{th}$ body, $\rr^\perp = (r_2,-r_1)$, and $\rho = \|\rr\|$.  Then,
any solution of the Stokes equations with a Dirichlet boundary condition
$\ff$ can be written in terms of the completed double-layer potential
\begin{align}
  \label{eqn:completed_DLP}
  \uu(\xx) = \DD[\eeta](\xx) + 
    \sum_{\ell=1}^{M} S[\llambda_\ell,\cc_\ell](\xx) +
    \sum_{\ell=1}^{M} R[\xi_\ell,\cc_\ell](\xx), \quad \xx \in \Omega,
\end{align}
where the density function, Stokeslets, and rotlets satisfy
\begin{subequations}
\label{eqn:completed_BIE}
\begin{align}
  \ff(\xx) &= -\frac{1}{2}\eeta(\xx) + \DD[\eeta](\xx) + 
      \sum_{\ell=1}^{M} S[\llambda_\ell,\cc_\ell](\xx) +
      \sum_{\ell=1}^{M} R[\xi_\ell,\cc_\ell](\xx) + \NN_0[\eeta](\xx),
      \quad \xx \in \bd\Omega, 
      \label{eqn:DLP}\\
  \llambda_\ell &= \frac{1}{2\pi}\int_{\gamma_\ell} \eeta(\yy) \, ds_\yy,
  \quad \ell=1,\ldots,M,
  \label{eqn:stokeslet} \\
  \xi_\ell &= \frac{1}{2\pi}\int_{\gamma_\ell} \yy^\perp \cdot \eeta(\yy)
  \, ds_\yy, \quad \ell=1,\ldots,M.
  \label{eqn:rotlet}
\end{align}
\end{subequations}
Here,
\begin{align*}
  \NN_0[\eeta](\xx) = \int_{\Gamma} 
    (\nn(\xx) \otimes \nn(\yy))\eeta(\yy) \, ds_\yy
\end{align*}
removes the rank one null space that results from the compatibility
constraint associated with the flux-free requirement of $\ff$.

@@

For numerical stability, we modify the erosion law \eqref{eqn:erosionModel} to include a term that depends on the local curvature $\kappa$. The idea is to prevent regions of very high
curvature from developing, as these regions would compromise the accuracy and stability of our fluid solver. We thus modify the normal velocity of each interface as follows
\begin{align}
  \label{eqn:Vn}
  \Vn = \atau + \eps \mean{\atau} \left( \frac{L }{2\pi} \kappa - 1
  \right).
\end{align}
The second term on the right is the curvature-dependent modification, the strength of which is controlled by the parameter $\eps \ll 1$. Here, $L$ is the total arc length of the body, and $\mean{\cdot}$ indicates a boundary mean. This term causes regions of high curvature to recede more rapidly, thus smoothing any high-frequency oscillations that my exist. Since a closed body has mean curvature $\mean{\kappa} = 2\pi / L$, this smoothing term has mean zero and and thus preserves area. Therefore, the only source of material loss is the shear-stress itself. In addition, we have scaled the smoothing term on the mean absolute shear, $\mean{\atau}$, so that the ratio of smoothing-to-erosion always remains fixed, even as bodies vanish.

@@

>>>>

Because of the length and velocity scales of groundwater flow, we
restrict our attention to small Reynolds number flows.  For groundwater
flow, Darcy's law is often used, but this does not resolve the pore
structure and assumes a homogeneous geometry which is often not valid.
Instead, we use the incompressible Stokes equations to represent the
flow, and the individual pores are resolved which results in a complex
multiply-connected geometry.  A boundary integral equation is
well-suited for solving the incompressible Stokes equations in a complex
geometry since it can resolve individual grains and the non-negligible
interactions between bodies are resolved. Moreover, high-order accuracy
and computational efficiency are obtained. To link the eroding body and
fluid dynamics, the body is eroded in the normal direction at a rate
that is proportional to the magnitude of the shear
stress~\cite{ris-moo-chi-she-zha2012}.  Tracking the interface of
fluid-structure interaction problems with a long time horizon is
particularly---the mesh is often distorted and tangled by the dynamics
and regions of high curvature develop.  These two issues are resolved
by, first, introducing a tangential velocity field to maintain a grid
that is equispaced in arc length, but that does not affect the dynamics,
and, second, a curvature penalization term.  Furthermore, the frame of
reference is changed to \thL~variables so that the curvature term
appears as a diffusive term that can easily be handled in a 
numerically stable fashion, e.g.~via an exponential integrator or an implicit method.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\paragraph{Contributions} The present study develops numerical methods
to simulate erosion of multiply-connected two-dimensional geometries.
Since groundwater flow occurs at small spatial and velocity scales, we
assume the flow is viscous (ie.~zero Reynolds number).  By allowing for
complex multiply-connected domains, we capture the effect that
interacting bodies have on erosion, and the grains in the flow are
individually resolved rather than homogenized.  We start by validating
our method on a single circular body placed symmetrically in a CHANNEL
flow.  By artificially fixing the area of the body as a function of
time, we are able to characterize the long-time behavior of the shape
and the resulting flow and compare these results with analytical
results.  Then, we compute the vorticity of erosion in
multiply-connected geometries to better characterize erosion in porous
media.

@@

A challenge with erosion is that it creates corners at points on the
body where the shear stress vanishes.  While recent work does allow for
integral equation methods to resolve corners, we employ a different
method.  
\todo[inline]{Maybe point out some drawbacks of using one of these BIE methods that allows corners. For example, are these methods more conceptually complex and/or more expensive? Maybe they are not quite at a mature stage for application? Another point: in the simulations we cannot predict where corners will form and so would have to identify them on the fly. Would this cause a problem if we tried to use a corner-resolving BIE?}
\todo[inline]{Could say our approach is simpler because  by smoothing corners, we can use the bare-bones BIE methods that are more well established compared to these newer, corner-resolving methods. Just some thoughts.}
First, we add a regularization term to the shear stress to
penalize regions of high curvature.  Second, we apply a Gaussian filter
to smooth the absolute value of the shear stress before the body shape
is updated.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\paragraph{Limitations}

\todo[inline]{These are some of my thoughts on reworking this first paragraph. See if you like it.}
We begin by making two fundamental design choices: first, we restrict ourselves to two spatial dimensions, and second the bodies are considered immobile (i.e.~fixed to their initial locations). The restriction to 2D allows us to simulate the erosion of many bodies (order 50-100) with high resolution, giving us the ability to accurately characterize the erosion of a porous medium. While BIE methods are available in 3D, the extra computational cost limits the number of bodies that could be practically simulated, making porous-medium simulations less feasible. The second assumption of immobile bodies affords a simple framework to study the effects of erosion in isolation. While other effects, such as sedimentation or transport, undoubtedly play a role in real applications, these effects would introduce additional complexity, both conceptually and numerically.

\todo[inline]{This is the original below:}
There are two main limitations of the current
work.  First, the flow is assumed to be two-dimensional.  However, by
assuming that the flow is two-dimensional, we are able to simulate the
erosion of more bodies and with higher resolution than would be
available in three dimensions.  Moreover, several flows can be described
to good accuracy under a two-dimensional approximation, but certain
phenomena occur only in three dimensions.  Extending the current work to
three dimensions is possible with a boundary integral equation
formulation.  In particular, the fluid solvers exist~\cite{mal-bir2015},
and the main challenges of a BIE solver have been addressed by other
groups including surface representation~\cite{yin-bir-zor2006,
vee-rah-bir-zor2011}, near-singular integration~\cite{yin-bir-zor2006,
kli-tor2016b}, and fast summation methods~\cite{yin-bir-zor2004,
ros-ols2015}.  The more significant challenge of extending our work to
three dimensions is guaranteeing that the mesh does not become
distorted, which can be done, for example, by using an extension of the
two-dimensional {\thL} formulation~\cite{amb-sie-tlu2013}.

@@

In this section, we describe the boundary integral equations employed for the fluid
equations \cite{Moore2007}
\todo[inline]{Bryan do you want to add a bunch of references to your boundary integral papers here (or somewhere else)?}
In particular, we solve the incompressible Stokes equations
with Dirichlet boundary conditions 
\begin{alignat*}{3}
  \mu\Delta \uu &= \grad p, \qquad && \xx \in \Omega, \\
  \grad \cdot \uu &= 0,   && \xx \in \Omega, \\
  \uu &= \ff,  && \xx \in \bd\Omega.
\end{alignat*}
The boundary condition $\ff$ is the prescribed velocity $\UU$ on
$\Gamma$, and it is zero on the grains $\gamma$.  As the viscosity $\mu$
also sets the time scale, we take $\mu=1$.  We start by defining the
double-layer potential
\todo[inline]{Do you think we should flesh out this time-scale stuff more precisely? i.e.~By setting $\mu = 1$ and $C = 1$ (the erosion constant), what exactly is the timescale that defines our dimensionless time.}
\begin{align}
  \DD[\eeta](\xx) = \frac{1}{\pi} \int_{\bd\Omega} 
    \frac{\rr \cdot \nn}{\rho^2} \frac{\rr \otimes \rr}{\rho^2} 
    \eeta(\yy) \, ds_\yy, \quad \xx \in \Omega,
    \label{eqn:velocityDLP}
\end{align}
where $\rr = \xx - \yy$, $\rho = \|\rr\|$, and $\eeta$ is an unknown density function.  The
double-layer potential cannot capture velocity fields corresponding to
rigid body motions~\cite{pow-mir1987}.  The rank deficiency is removed
by complementing the double-layer potential with a Stokeslets and
rotlets centered inside each interior body
\begin{align}
  S[\llambda_\ell,\cc_\ell] = \frac{1}{4\pi} \left(-\log \rho \II + 
    \frac{\rr \otimes \rr}{\rho^2} \right)\llambda_\ell
  \quad \text{and} \quad
  R[\xi_\ell,\cc_\ell] = \frac{\rr^\perp}{\rho^2}\xi_\ell, 
  \quad \ell=1,\ldots,M,
  \label{eqn:stokeslet_rotlet}
\end{align}
where  $\rr = \xx - \cc_\ell$, $\cc_\ell$ is a point inside the
$\ell^{th}$ body, $\rr^\perp = (r_2,-r_1)$, and $\rho = \|\rr\|$.  Then,
any solution of the Stokes equations with a Dirichlet boundary condition
$\ff$ can be written in terms of the completed double-layer potential
\begin{align}
  \label{eqn:completed_DLP}
  \uu(\xx) = \DD[\eeta](\xx) + 
    \sum_{\ell=1}^{M} S[\llambda_\ell,\cc_\ell](\xx) +
    \sum_{\ell=1}^{M} R[\xi_\ell,\cc_\ell](\xx), \quad \xx \in \Omega,
\end{align}
where the density function, Stokeslets, and rotlets satisfy
\begin{subequations}
\label{eqn:completed_BIE}
\begin{align}
  \ff(\xx) &= -\frac{1}{2}\eeta(\xx) + \DD[\eeta](\xx) + 
      \sum_{\ell=1}^{M} S[\llambda_\ell,\cc_\ell](\xx) +
      \sum_{\ell=1}^{M} R[\xi_\ell,\cc_\ell](\xx) + \NN_0[\eeta](\xx),
      \quad \xx \in \bd\Omega, 
      \label{eqn:DLP}\\
  \llambda_\ell &= \frac{1}{2\pi}\int_{\gamma_\ell} \eeta(\yy) \, ds_\yy,
  \quad \ell=1,\ldots,M,
  \label{eqn:stokeslet} \\
  \xi_\ell &= \frac{1}{2\pi}\int_{\gamma_\ell} \yy^\perp \cdot \eeta(\yy)
  \, ds_\yy, \quad \ell=1,\ldots,M.
  \label{eqn:rotlet}
\end{align}
\end{subequations}
Here,
\begin{align*}
  \NN_0[\eeta](\xx) = \int_{\Gamma} 
    (\nn(\xx) \otimes \nn(\yy))\eeta(\yy) \, ds_\yy
\end{align*}
removes the rank one null space that results from the compatibility
constraint associated with the flux-free requirement of $\ff$.

@@

For numerical stability, we modify the erosion law \eqref{eqn:erosionModel} to include a term that depends on the local curvature $\kappa$. The idea is to prevent regions of very high
curvature from developing, as these regions would compromise the accuracy and stability of our fluid solver. We thus modify the normal velocity of each interface as follows
\begin{align}
  \label{eqn:Vn}
  \Vn = \atau + \eps \mean{\atau} \left( \frac{L }{2\pi} \kappa - 1
  \right).
\end{align}
The second term on the right is the curvature-dependent modification, the strength of which is controlled by the parameter $\eps \ll 1$. Here, $L$ is the total arc length of the body, and $\mean{\cdot}$ indicates a boundary mean. This term causes regions of high curvature to recede more rapidly, thus smoothing any high-frequency oscillations that may exist. Since a closed body has mean curvature $\mean{\kappa} = 2\pi / L$, this smoothing term has mean zero and and thus preserves area. Therefore, the only source of material loss is the shear-stress itself. In addition, we have scaled the smoothing term on the mean absolute shear, $\mean{\atau}$, so that the ratio of smoothing-to-erosion always remains fixed, even as bodies vanish.

@@

\todo[inline]{Mention key term exponential integrator for stiff term.}
